\section{Injective and Projective Resolutions}


\begin{definition}
  \label{definition of projective and injective}
  Let~$\Ccat$ be a category.
  \begin{enumerate}
    \item
      An object~$P \in \Ob(\Ccat)$ is \emph{projective}\index{projective!object}\index{object!projective} if for every epimorphism~$f \colon X \to Y$ in~$\Ccat$ and every morphism~$g \colon P \to Y$ in~$\Ccat$ there exists a lift of~$g$ along~$f$, i.e.\ a morphism~$g' \colon P \to X$ that makes the following diagram commute:
      \[
        \begin{tikzcd}
            {}
          & P
            \arrow{d}[right]{g}
            \arrow[dashed]{dl}[above left]{g'}
          \\
            X
            \arrow{r}[below]{f}
          & Y
        \end{tikzcd}
      \]
    \item
      An object~$I \in \Ob(\Ccat)$ is \emph{injective}\index{injective!object}\index{object!injective} if for every monomorphism~$f \colon X \to Y$ and every morphism~$g \colon X \to I$ there exist a morphism~$g' \colon Y \to I$ that makes the following diagram commute:
      \[
        \begin{tikzcd}
            X
            \arrow{r}[above]{f}
            \arrow{d}[left]{g}
          & Y
            \arrow[dashed]{dl}[below right]{g'}
          \\
            I
          & {}
        \end{tikzcd}
      \]
  \end{enumerate}
\end{definition}


\begin{remark}
  Let~$\Ccat$ be a category.
  \begin{enumerate}
    \item
      An object~$X \in \Ob(\Ccat)$ is projective in~$\Ccat$ if and only if it is injective in~$\Ccat^\op$.
    \item
      The induced morphisms~$g'$ in \cref{definition of projective and injective} are in general not unique.
  \end{enumerate}
\end{remark}


\begin{lemma}
  \leavevmode
  \begin{enumerate}
    \item
      \label{characterization of projectives}
      For an object~$P \in \Ob(\Acat)$ the following conditions are equivalent:
      \begin{enumerate}
        \item
          \label{is projective}
          The object~$P$ is projective.
        \item
          \label{is exact}
          The functor~$\Hom_{\Acat}(P,-) \colon \Acat \to \Ab$ is exact.
        \item
          \label{sends epis to epis}
          The functor~$\Hom_{\Acat}(P,-) \colon \Acat \to \Ab$ maps epimorphisms to epimorphisms.
      \end{enumerate}
    \item
      For an object~$I \in \Ob(\Acat)$ the following conditions are equivalent:
      \begin{enumerate}
        \label{characterization of injectives}
        \item
          The object~$I$ is injective.
        \item
          The functor~$\Hom_\Acat(-,I) \colon \Acat^\op \to \Ab$ is exact.
        \item
          The functor~$\Hom_\Acat(-,I) \colon \Acat^\op \to \Ab$ maps epimorphisms to epimorphisms.%
          \footnote{If one thinks about~$\Hom_\Acat(-,I)$ not as a covariant functor~$\Acat^\op \to \Ab$ but instead as a contravariant functor~$\Acat \to \Ab$, then this means that~$\Hom_{\Acat}(-,I)$ maps monomorphisms (in~$\Acat$) to epimorphisms.}
      \end{enumerate}
  \end{enumerate}
\end{lemma}


\begin{proof}
  It sufficies to show part~\ref*{characterization of projectives} because part~\ref*{characterization of injectives} then follows by duality.
  
  \begin{description}
    \item[\ref*{is projective}~$\iff$~\ref*{sends epis to epis}]
      That the functor~$\Hom_{\Acat}(P,-)$ maps epimorphisms to surjections is just a reformulation of the definition of a projective object.
    \item[\ref*{sends epis to epis}~$\implies$~\ref*{is exact}]
      This holds because the functor~$\Hom_\Acat(P,-)$ is already left exact.
    \item[\ref*{is exact}~$\implies$~\ref*{sends epis to epis}]
      Every epimorphism~$f \colon X \to Y$ can be extended to a short exact sequence
      \[
        0
        \to
        \ker(f)
        \to
        X
        \to
        Y
        \to
        0
      \]
      in~$\Acat$, and the exactness of the induced sequence
      \[
        0
        \to
        \Hom_\Acat(P,\ker(f))
        \to
        \Hom_\Acat(P,X)
        \to
        \Hom_\Acat(P,Y)
        \to
        0
      \]
      entails that the group homomorphism~$\Hom_\Acat(P,X) \to \Hom_\Acat(P,Y)$ is surjective.
    \qedhere
  \end{description}
\end{proof}


% TODO: Projective modules.
% TODO: Remarks about injective modules.


\begin{definition*}
  Let~$\Ccat$ be a category and let~$s \colon X \to Y$ and~$r \colon Y \to X$ be two morphisms in~$\Ccat$ with~$rs = \id_X$.
  Then the morphism~$s$ is a \emph{section \textup(for~$r$\textup)}\index{section} and the morphism~$r$ is a \emph{retraction \textup(for~$s$\textup)}\index{retraction}.
  (This definition is given in Exercise~2 of Exercise~Sheet~9.)
\end{definition*}


\begin{remark*}
  Let~$\Ccat$ and~$\Dcat$ be categories.
  \begin{enumerate}
    \item
      Sections are also called \emph{split monomorphisms}\index{split!monomorphism}\index{monomorphism!split}, and retractions are also called \emph{split epimorphisms}\index{split!epimorphism}\index{epimorphism!split}.
  
      Split monomorphisms are monomorphisms, and split epimorphisms are epimorphisms (as the names indicate).
      (This observation is also made in Exercise~2 of Exercise~Sheet~9.)
      Indeed, if~$s \colon X \to Y$ and~$r \colon Y \to X$ are morphisms in~$\Ccat$ with~$rs = \id_X$ then it follows for all morphisms~$t, t' \colon W \to X$ that
      \[
        st = st'
        \implies
        rst = rst
        \implies
        t = t' \,,
      \]
      and simiarly for all morphisms~$u, u' \colon Y \to Z$ that
      \[
        ur = u'r
        \implies
        urs = u'rs
        \implies
        u = u'  \,.
      \]
      
      A monomorphism is said to be \emph{split} if it is a split monomorphism.
      Similary, an epimorphism is said to be \emph{split} if it is a split epimorphism.
    \item
      A morphism~$f$ in~$\Ccat$ is a section in~$\Ccat$ if and only if it is a retraction in~$\Ccat^\op$.
    \item
      A morphism~$f$ in~$\Ccat$ is both a section and a rectration if and only it is an isomorphism.
    \item
      Functors respect sections and retractions:
      Let~$F \colon \Ccat \to \Dcat$ be a functor, let~$s$ is a section and~$\Ccat$ and~$r$ is a retraction in~$\Ccat$.
      Then~$F(s)$ is a section in~$\Dcat$ and~$F(r)$ is a retraction in~$\Dcat$.
  \end{enumerate}
\end{remark*}


\begin{lemma*}
  \label{characterizations of split ses}
  For a short exact sequence
  \[
    0
    \to
    X'
    \xto{f}
    X
    \xto{g}
    X''
    \to
    0
  \]
  in~$\Acat$, the following conditions are equivalent:
  \begin{enumerate}
    \item
      \label{is left split}
      The morphism~$f$ is a section.
    \item
      \label{is right split}
      The morphism~$g$ is a retraction.
    \item
      \label{is total split}
      There exists an isomorphism~$\alpha \colon X \to X' \oplus X''$ that makes the diagram
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & X'
            \arrow{r}[above]{f}
            \arrow[equal]{d}
          & X
            \arrow{r}[above]{g}
            \arrow[dashed]{d}[right]{\alpha}
          & X''
            \arrow{r}
            \arrow[equal]{d}
          & 0
          \\
            0
            \arrow{r}
          & X'
            \arrow{r}
          & X' \oplus X''
            \arrow{r}
          & X''
            \arrow{r}
          & 0
        \end{tikzcd}
      \]
      commute, where~$X' \to X' \oplus X''$ and~$X' \oplus X'' \to X''$ are the canonical morphisms that are part of the biproduct structure of~$X' \oplus X''$.
    \item
      \label{is biproduct}
      There exists morphisms~$r \colon X \to X'$ and~$s \colon X'' \to X$ such that~$(X, (f,s), (r,g))$ is a biproduct of~$X'$ and~$X''$.
  \end{enumerate}
\end{lemma*}


\begin{proof*}
  The equivalence of the parts~\ref*{is left split},~\ref*{is right split} and~\ref*{is biproduct} is Exercise~2 of Exercise~sheet~9.
  
  The implication \ref*{is total split}~$\implies$~\ref*{is biproduct} can be seen by pulling back the~$X'$\nobreakdash-$X''$\nobreakdash-biproduct structure of~$X' \oplus X''$ along~$\alpha$ to a~$X'$\nobreakdash-$X''$\nobreakdash-biproduct structure on~$X$.
  It then follows from the commutativity of the given diagram that the canonical morphism~$X' \to X' \oplus X''$ corresponds to the morphism~$f \colon X' \to X$, and that the canonical morphism~$X' \oplus X'' \to X''$ corresponds to the morphism~$g \colon X \to X''$.
  
  For the implications \ref*{is left split}~$\implies$~\ref*{is total split} one chooses a retraction~$r \colon X \to X'$ of~$f$.
  Then the morphism
  \[
    \alpha
    \defined
    \begin{bmatrix}
      r \\
      g
    \end{bmatrix}
    \colon
    X
    \to
    X' \oplus X''
  \]
  makes the given diagram commute, and it follows from the \hyperref[5 lemma]{\dash{$5$}{lemma}} that~$\alpha$ is an isomorphism.
\end{proof*}


% TODO: Make this a proper proof.


\begin{definition*}
  A short exact sequence in~$\Acat$ is \emph{split}\index{split!short exact sequence}\index{short exact sequence!split} if it satisfies the equivalent conditions from~\cref{characterizations of split ses}.
\end{definition*}


\begin{warning*}
  While it still makes sense to talk about short exact sequences in the category~$\Group$ of groups, it is for such a short exact sequence of groups
  \[
    1
    \to
    K
    \xlongto{f}
    G
    \xlongto{g}
    H
    \to
    1
  \]
  not equivalent that~$f$ is a section and that~$g$ is a retraction.
  Indeed, that~$f$ is a section is---roughly speaking---equivalent to~$G$ being a direct product of the groups~$K$ and~$H$;
  whereas~$g$ being a retract is equivalent to~$G$ being a semidirect product of the groups~$K$ and~$H$.
\end{warning*}


\begin{remark*}[Split chain complexes]
  \leavevmode
  \begin{enumerate}
    \item
      If more generally~$\Ccc$ is a chain complex in~$\Acat$, then~$\Ccc$ \emph{splits}\index{split!chain complex}\index{chain complex!split} if there exists a family~$s = (s_n)_{n \in \Integer}$ of morphisms~$s_n \colon C_n \to C_{n+1}$ with~$dsd = d$, i.e.\ such that~$d_n s_n d_n = d_n$ for every~$n \in \Integer$.
      Such a family~$s$ is a \emph{split} for~$\Ccc$, and may be visualized as follows:
      \[
        \begin{tikzcd}[column sep = large]
            \dotsb
            \arrow{r}
          & C_{n+1}
            \arrow[dashed, bend right=55]{l}
            \arrow{r}[above]{d_{n+1}}
          & C_n
            \arrow[dashed, bend right=55]{l}[above]{s_n}
            \arrow{r}[above]{d_n}
          & C_{n-1}
            \arrow[dashed, bend right=55]{l}[above]{s_{n-1}}
            \arrow{r}
          & \dotsb
            \arrow[dashed, bend right=55]{l}
        \end{tikzcd}
      \]
    \item
      A more intuitive but equivalent definition of~$\Ccc$ being split is the following:
      There exist families~$(B_n)_{n \in \Integer}$ and~$(H_n)_{n \in \Integer}$ of objects~$B_n, H_n \in \Ob(\Acat)$, and isomorphisms
      \[
        \alpha_n
        \colon
        C_n
        \to
        B_n \oplus H_n \oplus B_{n-1}
      \]
      that make the following diagram commute for every~$n \in \Integer$:
      \[
        \begin{tikzcd}[ampersand replacement = \&, column sep = 4em]
              \dotsb
              \arrow{r}
          \&  C_n
              \arrow{r}[above]{d_n}
              \arrow{d}[right]{\alpha_n}
          \&  C_{n-1}
              \arrow{r}
              \arrow{d}[right]{\alpha_{n-1}}
          \&  \dotsb
          \\
              \dotsb
              \arrow{r}
          \&  B_n \oplus H_n \oplus B_{n-1}
              \arrow{r}[below]{\begin{bsmallmatrix} 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 0 & 0 \end{bsmallmatrix}}
          \&  B_{n-1} \oplus H_{n-1} \oplus B_{n-2}
              \arrow{r}
          \&  \dotsb
        \end{tikzcd}
      \]
      (This characterization of split chain complexes should be compared with the characterization~\ref*{is total split} of a split short exact sequence from \cref{characterizations of split ses}.)
      The split morphisms~$s_n \colon C_n \to C_{n+1}$ correspond to the morphisms
      \[
        \begin{bmatrix}
          0 & 0 & 0 \\
          0 & 0 & 0 \\
          1 & 0 & 0
        \end{bmatrix}
        \colon
        B_n \oplus H_n \oplus B_{n-1}
        \to
        B_{n+1} \oplus H_{n+1} \oplus B_n \,.
      \]
      We furthermore have that~$\Bl_n(\Ccc) \cong B_n$,~$\Zl_n(\Ccc) \cong B_n \oplus H_n$ and~$\Hl_n(\Ccc) \cong H_n$ for every~$n \in \Integer$.
      Moverover, under these identifications the canonical (mono)mor\-phism~$\Bl_n(\Ccc) \to \Zl_n(\Ccc)$ corresponds to the canonical morphism~$B_n \to B_n \oplus H_n$, and the canonical (epi)morphism~$\Zl_n(\Ccc) \to \Hl_n(\Ccc)$ corresponds to the canonical morphism~$B_n \oplus H_n \to H_n$.
    \item
      A short exact sequence
      \[
        0
        \to
        X'
        \xlongto{f}
        X
        \xlongto{g}
        X''
        \to
        0
      \]
      in~$\Acat$ can be regarded as a chain complex
      \[
        \dotsb
        \to
        0
        \to
        0
        \to
        X'
        \xlongto{f}
        X
        \xlongto{g}
        X''
        \to
        0
        \to
        0
        \to
        \dotsb
      \]
      in~$\Acat$.
      Then both notions of \enquote{being split} coincide:
      
      The chain complex is split (according to the above definition) if and only if there exists morphisms~$r \colon X \to X'$ and~$s \colon X'' \to X$ such that~$frf = f$ and~$gsg = g$.
      \[
        \begin{tikzcd}[column sep = 3.2em]
            \dotsb
            \arrow{r}
          & 0
            \arrow{r}
          & X'
            \arrow{r}[above]{f}
          & X
            \arrow[dashed, bend right=55]{l}[above]{r}
            \arrow{r}[above]{g}
          & X''
            \arrow[dashed, bend right=55]{l}[above]{s}
            \arrow{r}
          & 0
            \arrow{r}
          & \dotsb
        \end{tikzcd}
      \]
      That~$frf = f$ is equivalent to~$rf = \id_{X'}$ because~$f$ is a monomorphism, and that~$gsg = g$ is equivalent to~$gs = \id_{X''}$ because~$g$ is an epimorphism.
  \end{enumerate}
\end{remark*}


\begin{definition*}
  Let~$X \in \Ob(\Acat)$ be an object in~$\Acat$.
  An object~$Y \in \Ob(\Acat)$ is a \emph{direct summand}\index{direct summand} of~$X$ if there exists another object~$Y' \in \Ob(\Acat)$ with~$X \cong Y \oplus Y'$.
\end{definition*}


\begin{lemma}
  Let~$P, I \in \Ob(\Acat)$ be a objects in~$\Acat$.
  \begin{enumerate}
    \item
      \label{ses ending in projective}
      The object~$P$ is projective if and only if every short exact sequence of the form~$0 \to X' \to X \to P \to 0$ in~$\Acat$ splits.
    \item
      Dually, the object~$I$ is injective if and only if every short exact sequence of the form~$0 \to I \to X \to X'' \to 0$ in~$\Acat$ splits.
    \item
      \label{direct summand of projective}
      If~$P$ is projective then every direct summand of~$P$ is again projective.
    \item
      If~$I$ is injective then every direct summand of~$I$ is again injective.
  \end{enumerate}
\end{lemma}


\begin{proof}
  \leavevmode
  \begin{enumerate}
    \item
      Suppose that~$P$ is projective and let~$0 \to X' \to X \to P \to 0$ be a short exact sequence in~$\Acat$ that ends in~$P$.
      Then there exists a lift of the identity~$\id_P \colon P \to P$ along the epimorphism~$X \to P$, that is then a section for~$X \to P$.
      Hence~$X \to P$ is a retraction, which shows that the given short exact sequence splits.
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & X'
            \arrow{r}
          & X
            \arrow{r}
          & P
            \arrow{r}
          & 0
          \\
            {}
          & {}
          & {}
          & P
            \arrow{u}[right]{\id_P}
            \arrow[dashed]{ul}
          & {}
        \end{tikzcd}
      \]
      
      Suppose now on the other hand that every short exact sequence that ends in~$P$ splits.
      Let~$f \colon X \to Y$ be an epimorphism in~$\Acat$ and let~$g \colon P \to Y$ be a morphism in~$\Acat$.
      We may extend the epimorphism~$f$ to a short exact sequence
      \[
        0
        \to
        \ker(f)
        \to
        X
        \xlongto{f}
        Y
        \to
        0
      \]
      in~$\Acat$.
      By using \cref{kernels of pullbacks} we get a commutative diagram
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & \ker(f)
            \arrow{r}
            \arrow[equal]{d}
          & X'
            \arrow{r}[above]{f'}
            \arrow{d}[right]{g'}
          & P
            \arrow{r}
            \arrow{d}[right]{g}
          & 0
          \\
            0
            \arrow{r}
          & \ker(f)
            \arrow{r}
          & X
            \arrow{r}[above]{f}
          & Y
            \arrow{r}
          & 0
        \end{tikzcd}
      \]
      with exact rows.
      The upper row splits by assumption, hence there exists a section~$s \colon P \to X'$ for~$f'$.
      For the morphism~$g'' \defined g's \colon P \to X$ we then have
      \[
          f g''
        = f g' s
        = g f' s
        = g \id_P
        = g \,.
      \]
    \item
      This is dual to part~\ref*{ses ending in projective}.
    \item
      This is left as an exercise.
%     TODO: Add a reference to the exercise sheet.
    \item
      This is dual to part~\ref*{direct summand of projective}.
    \qedhere
  \end{enumerate}
\end{proof}


\begin{remark*}
  Let~$\Ccat$ be a category.
  \begin{enumerate}
    \item
      If~$(P_j)_{j \in J}$ is a family of projective objects~$P_j$ in~$\Ccat$ that admit a coproduct~$\coprod_{j \in J} P_j$ then~$\coprod_{j \in J} P_j$ is again projective.
      
      If there exists in the category~$\Ccat$ for any two objects~$X, Y \in \Ob(\Ccat)$ a morphism~$X \to Y$ (e.g.\ if~$\Ccat$ is preadditive or has a zero object) then the converse also holds:
      If~$\coprod_{j \in J} P_j$ is projective then~$P_j$ is projective for every~$j \in J$.
%     TODO: Prove this.
    \item
      If~$(I_j)_{j \in J}$ is a family of injective objects~$I_j$ in~$\Ccat$ that admit a product~$\prod_{j \in J} I_j$ then~$\prod_{j \in J} I_j$ is again injective.
      
      If there exists in the category~$\Ccat$ for any two objects~$X, Y \in \Ob(\Ccat)$ a morphism~$X \to Y$ then the converse also holds:
      If~$\prod_{j \in J} I_j$ is injective then~$I_j$ is injective for every~$j \in J$.
  \end{enumerate}
\end{remark*}


\begin{definition}
  \leavevmode
  \begin{enumerate}
    \item 
      The category~$\Acat$ \emph{has enough projectives}\index{enough!projectives} if for every object~$X \in \Ob(\Acat)$ there exists an epimorphism~$P \to X$ coming from some projective object~$P \in \Ob(\Acat)$.
    \item
      The category~$\Acat$ \emph{has enough injectives}\index{enough!injectives} if for every object~$X \in \Ob(\Acat)$ there exists a monomorphism~$X \to I$ into some injective object~$I \in \Ob(\Acat)$.
  \end{enumerate}
\end{definition}


\begin{definition}
  Let~$X \in \Ob(\Acat)$ be an object in~$\Acat$.
  \begin{enumerate}
    \item
      A \emph{chain resolution}\index{chain!resolution}\index{resolution!chain} of~$X$ is a pair~$(\Ccc, p_0)$ consisting of
      \begin{itemize}
        \item
          a chain complex~$\Ccc \in \Chh_{\geq 0}(\Acat)$ bounded below by degree~$0$, together with
        \item
          a morphism~$p_0 \colon C_0 \to X$,
      \end{itemize}
      such that the resulting sequence
      \[
        \dotsb
        \to
        C_2
        \xlongto{d_2}
        C_1
        \xlongto{d_1}
        C_0
        \xlongto{p_0}
        X
        \to
        0
      \]
      is exact.
    \item
      A \emph{projective resolution}\index{projective!resolution}\index{resolution!projective} of~$X$ is a chain resolution~$(\Pcc, p_0)$ of~$X$ in which~$P_n$ is projective for every~$n \geq 0$.
    \item
      A \emph{cochain resolution}\index{cochain!resolution}\index{resolution!cochain} of~$X$ is a pair~$(\Cccc, i^0)$ consisting of
      \begin{itemize}
        \item
          a cochain complex~$\Cccc \in \CChh^{\geq 0}(\Acat)$ bounded below by degree~$0$, together with
        \item
          a morphism~$i^0 \colon X \to C^0$,
      \end{itemize}
      such that the resulting sequence
      \[
        0
        \to
        X
        \xlongto{i^0}
        C_0
        \xlongto{d^0}
        C_1
        \xlongto{d^1}
        C_2
        \to
        \dotsb
      \]
      is exact.
    \item
      An \emph{injective resolution}\index{injective!resolution}\index{resolution!injective} of~$X$ is a cochain resolution~$(\Iccc, i^0)$ of~$X$ in which~$I_n$ is injective for every~$n \geq 0$.
  \end{enumerate}
\end{definition}


\begin{remark}
  Let~$X \in \Ob(\Acat)$ be an object in~$\Acat$.
  \begin{enumerate}
    \item
      We can consider the object~$X$ as a chain (resp.\ cochain) complex that is concentrated in degree~$0$.
      Then~$\Hl_0(X) = X$ (resp.\ $\Hl^0(X) = X$) and~$\Hl_n(X) = 0$ (resp.\ $\Hl^n(X) = 0$) for every~$n \neq 0$.
    \item
      For a chain complex~$\Ccc \in \Ch(\Acat)$, a morphism~$f \colon \Ccc \to X$ is uniquely determined by the single morphism~$C_0 \to X$ (for all~$n \neq 0$ it holds that~$f_n = 0$), and the morphism~$f_0$ is subject to the single conditions~$f_0 d_1 = 0$.
      Indeed, this is precisely what it means for the following diagram to commute:
      \[
        \begin{tikzcd}
            \dotsb
            \arrow{r}
          & C_2
            \arrow{r}[above]{d_2}
            \arrow{d}
          & C_1
            \arrow{r}[above]{d_1}
            \arrow{d}
          & C_0
            \arrow{r}[above]{d_0}
            \arrow{d}[right]{f_0}
          & C_{-1}
            \arrow{r}[above]{d_{-1}}
            \arrow{d}
          & C_{-2}
            \arrow{r}
            \arrow{d}
          & \dotsb
          \\
            \dotsb
            \arrow{r}
          & 0
            \arrow{r}
          & 0
            \arrow{r}
          & X
            \arrow{r}
          & 0
            \arrow{r}
          & 0
            \arrow{r}
          & \dotsb
        \end{tikzcd}
      \]
    \item
      For a cochain complex~$\Cccc \in \CCh(\Acat)$, a morphism~$f \colon X \to \Cccc$ is uniquely determined by the single morphism~$f^0 \colon X \to C^0$ (for all~$n \neq 0$ it holds that~$f^n = 0$), and the morphism~$f^0$ is subject to the single condition~$d^0 f^0 = 0$.
      Indeed, this is precisely what it means for the following diagram to commute:
      \[
        \begin{tikzcd}
            \dotsb
            \arrow{r}
          & 0
            \arrow{r}
            \arrow{d}
          & 0
            \arrow{r}
            \arrow{d}
          & X
            \arrow{r}
            \arrow{d}[right]{f^0}
          & 0
            \arrow{r}
            \arrow{d}
          & 0
            \arrow{r}
            \arrow{d}
          & \dotsb
          \\
            \dotsb
            \arrow{r}
          & C^{-2}
            \arrow{r}[above]{d^{-2}}
          & C^{-1}
            \arrow{r}[above]{d^{-1}}
          & C^0
            \arrow{r}[above]{d^0}
          & C^1
            \arrow{r}[above]{d^1}
          & C^2
            \arrow{r}
          & \dotsb
        \end{tikzcd}
      \]
  \end{enumerate}
\end{remark}


\begin{lemma}
  Let~$X \in \Ob(\Acat)$ be an object in~$\Acat$.
  \begin{enumerate}
    \item
      \label{resolution iff qim}
      Let~$\Ccc \in \Chh_{\geq 0}(\Acat)$ be a chain complex that is bounded below by degree~$0$, and let~$p_0 \colon C_0 \to X$ be a morphism with~$p_0 d_1 = 0$.
      Let~$p \colon \Ccc \to X$ be the corresponding morphism of chain complexes.
      Then the pair~$(\Ccc, p_0)$ is a chain resolution of~$X$ if and only if the morphism of chain complexes~$p$ is a {\qim}.
    \item
      \label{coresolution iff qim}
      Let~$\Cccc \in \CChh^{\geq 0}(\Acat)$ be a cochain complex this is bounded below by degree~$0$, and let~$i^0 \colon X \to C^0$ be a morphism with~$d^0 i^0 = 0$.
      Let~$i \colon X \to \Cccc$ be the corresponding morphism of cochain complexes.
      Then the pair~$(\Cccc, i^0)$ is a cochain resolution of~$X$ if and only if the morphism of cochain complexes~$i$ is a {\qim}.
  \end{enumerate}
\end{lemma}


\begin{proof}
  It sufficies to prove part~\ref*{resolution iff qim}, as part~\ref*{coresolution iff qim} follows by duality.
  
  That~$(\Ccc, p_0)$ is a chain resolution of~$X$, i.e.\ that the sequence
  \[
    \dotsb
    \to
    C_2
    \xlongto{d_2}
    C_1
    \xlongto{d_1}
    C_0
    \xlongto{p_0}
    X
    \to
    0
  \]
  is exact, means that~$\Hl_n(\Ccc) = 0$ for every~$n \neq 0$, that~$C_1 \xto{d_1} C_0 \xto{p_0} X \to 0$ is exact.
  That the morphism~$p \colon \Ccc \to X$ given by the commutative diagram
  \[
    \begin{tikzcd}
        \dotsb
        \arrow{r}
      & C_2
        \arrow{r}[above]{d_2}
        \arrow{d}
      & C_1
        \arrow{r}[above]{d_1}
        \arrow{d}
      & C_0
        \arrow{r}
        \arrow{d}[right]{p_0}
      & 0
        \arrow{r}
        \arrow{d}
      & 0
        \arrow{r}
        \arrow{d}
      & \dotsb
      \\
        \dotsb
        \arrow{r}
      & 0
        \arrow{r}
      & 0
        \arrow{r}
      & X
        \arrow{r}
      & 0
        \arrow{r}
      & 0
        \arrow{r}
      & \dotsb
    \end{tikzcd}
  \]
  is a {\qim} means that~$\Hl_n(\Ccc) = 0$ for every~$n \neq 0$ and that the induced morphism~$\coker(d_1) \to X$ is an isomorphism.
  This induced morphism is the unique morphism that makes the triangle
  \[
    \begin{tikzcd}
        C_0
        \arrow{r}
        \arrow{dr}[below left]{p_0}
      & \coker(d_1)
        \arrow[dashed]{d}
      \\
        {}
      & X
    \end{tikzcd}
  \]
  commute.
  The commutativity of this triangle together with~$\coker(d_1) \to X$ being an isomorphism is equivalent to~$p_0$ being a cokernel of~$d_1$, which is in turn equivalent to the exactness of the sequence~$C_1 \xto{d_1} C_0 \xto{p_0} X \to 0$.
\end{proof}


\begin{lemma}
  \leavevmode
  \begin{enumerate}
    \item
      \label{existence of projective resolution}
      If the category~$\Acat$ has enough projectives, then every object~$X \in \Ob(\Acat)$ admits a projective resolution.
    \item
      \label{existence of injective resolution}
      If the category~$\Acat$ has enough injectives, then every object~$X \in \Ob(\Acat)$ admits an injective resolution.
  \end{enumerate}
\end{lemma}


\begin{proof}
  It sufficies to prove part~\ref*{existence of projective resolution}, as part~\ref*{existence of injective resolution} follows from part~\ref*{existence of projective resolution} by duality.
  So suppose that the abelian category~$\Acat$ has enough enough projectives.
  
  For any object~$X \in \Ob(\Acat)$ there exists by assumption an epimorphism~$p_0 \colon P_0 \to X$ for some projective object~$P_0$ in~$\Acat$.
  We have thus an exact sequence.
  \[
    P_0
    \xlongto{p_0}
    X
    \to
    0 \,.
  \]
  Suppose now that we have already constructed an exact sequence
  \[
    P_n
    \xlongto{d_n}
    P_{n-1}
    \xlongto{d_{n-1}}
    \dotsb
    \xlongto{d_1}
    P_0
    \xlongto{p_0}
    X
    \to
    0 \,.
  \]
  for some~$n \geq 0$.
  For the kernel~$\ker(d_n)$ there exists an epimorphism
  \[
    \tilde{d}_{n+1}
    \colon
    P_{n+1}
    \to
    \ker(d_n)
  \]
  for some projective object~$P_{n+1}$ in~$\Acat$.
  It follows for the composition
  \[
    d_{n+1}
    \colon
    P_{n+1}
    \xlongto{\tilde{d}_{n+1}}
    \ker(d_n)
    \to
    P_n
  \]
  from the exactness of the sequence
  \[
    \ker(d_n)
    \to
    P_n
    \xlongto{d_n}
    P_{n-1}
  \]
  that the sequence
  \[
    P_{n+1}
    \xlongto{d_{n+1}}
    P_n
    \xlongto{d_n}
    P_{n-1}
  \]
  is again exact, as seen in~\cref{composing exact sequence with epi and mono}.
  We hence arrive at the exact sequence
  \[
    P_{n+1}
    \xlongto{d_{n+1}}
    P_n
    \xlongto{d_n}
    P_{n-1}
    \xlongto{d_{n-1}}
    \dotsb
    \xlongto{d_1}
    P_0
    \xlongto{p_0}
    X
    \to
    0 \,.
  \]
  By induction we get the desired projective resolution of~$X$.
%   Let~$d_{n+1} \colon P_{n+1} \to P_n$ be the composition
%   \[
%     d_{n+1}
%     \colon
%     P_{n+1}
%     \xlongto{\tilde{d}_{n+1}}
%     \ker(d_n)
%     \to
%     P_n \,.
%   \]
%   Then
%   \[
%       \im(d_{n+1})
%     = \im(\ker(d_n) \to P_n)
%     = \ker(d_n) \,,
%   \]
%   where the first equality holds because~$\tilde{d}_{n+1}$ is an epimorphism, and the second equality because every monomorphism is the kernel of its cokernel.
\end{proof}


\begin{example}
  We give a short overview about which abelian categories have enough projectives or enough injectives.
  (Details on this will be given in the upcoming Chapter~6.)
  \begin{center}
    \begingroup
    \renewcommand{\arraystretch}{2}
    \begin{tabular}{ccc}
        category
      & enough projectives
      & enough injectives
      \\
      \hline
        \begingroup
        \renewcommand{\arraystretch}{1}
        \begin{tabular}{c}
          $\Modl{A}$ and~$\Modr{A}$ \\
          where~$A$ is a~{\kalg}
        \end{tabular}
        \endgroup
      & Yes
      & Yes
      \\
        $\Sheaf_X(\Ab)$
      & No (in general)
      & Yes
      \\
        $\Presheaf_X(\Ab)$
      & Yes
      & Yes
      \\
        $\Modlfg{\Integer}$
      & Yes
      & No
      \\
        \begingroup
        \renewcommand{\arraystretch}{1}
        \begin{tabular}{c}
          $\Modlfd{A}$ and~$\Modrfd{A}$ \\
          where~$\kf$ is a field  \\
          and~$A$ is a f.d.~{\kalg}
        \end{tabular}
        \endgroup
      & Yes
      & Yes
    \end{tabular}
    \endgroup
  \end{center}
\end{example}





\lecturend{19}





\begin{theorem}[Comparison theorem]\index{comparison theorem}\index{theorem!comparison}
  Let~$f \colon X \to Y$ be a morphism in~$\Acat$.
  \begin{enumerate}
    \item
      \label{comparison for projective}
      Let~$\Pcc \xto{p_0} X$ be a projective resolution of~$X$ and let~$\Ccc \xto{c_0} Y$ be a chain resolution of~$Y$.
      Then there exists a morphism of chain complexes~$\hat{f} \colon \Pcc \to \Ccc$ that makes the square
      \begin{equation}
        \label{comparison square for projective}
        \begin{tikzcd}
            P_0
            \arrow{r}[above]{p_0}
            \arrow[dashed]{d}[left]{\hat{f}_0}
          & X
            \arrow{d}[right]{f}
          \\
            C_0
            \arrow{r}[above]{c_0}
          & Y
        \end{tikzcd}
      \end{equation}
      commute.
      The morphism~$\hat{f}$ is uniquely determined up to homotopy by this property.
    \item
      \label{comparison for injective}
      Let~$Y \xto{i^0} \Iccc$ be an injective resolution of~$Y$ and let~$X \xto{c^0} \Cccc$ be a cochain resolution of~$X$.
      Then there exists a morphism of cochain complexes~$\hat{f} \colon \Cccc \to \Iccc$ that makes the square
      \begin{equation}
        \label{comparison square for injective}
        \begin{tikzcd}
            X
            \arrow{r}[above]{c^0}
            \arrow{d}[left]{f}
          & C^0
            \arrow[dashed]{d}[right]{\hat{f}^0}
          \\
            Y
            \arrow{r}[above]{i^0}
          & I^0
        \end{tikzcd}
      \end{equation}
      commute.
      The morphism~$\hat{f}$ is uniquely determined up to homotopy by this property.
  \end{enumerate}
\end{theorem}


\begin{proof}
  It sufficies to prove part~\ref*{comparison for projective}, as part~\ref*{comparison for injective} follows from part~\ref*{comparison for projective} by duality.
  We first construct the morphism~$\hat{f}$, and then show that it is unique up to homotopy.
  
  We set for~$P_{-1} \defined X$ and~$P_{-2} \defined 0$ with~$d^P_0 \defined p_0$ and~$d^P_{-1} = 0$, and we similarly set~$C_{-1} \defined Y$ and~$C_{-2} \defined 0$ with~$d^C_0 = c_0$ and~$d^C_{-1} = 0$.
  We also set $\hat{f}_{-1} \defined f$ and~$\hat{f}_{-2} = 0$.
  The given commutative diagram
  \[
    \begin{tikzcd}
        \dotsb
        \arrow{r}
      & P_2
        \arrow{r}[above]{d_2}
      & P_1
        \arrow{r}[above]{d_1}
      & P_0
        \arrow{r}[above]{p_0}
      & X
        \arrow{r}
        \arrow{d}[right]{f}
      & 0
        \arrow{d}
      \\
        \dotsb
        \arrow{r}
      & C_2
        \arrow{r}[below]{d_2}
      & C_1
        \arrow{r}[below]{d_1}
      & C_0
        \arrow{r}[below]{c_0}
      & Y
        \arrow{r}
      & 0
    \end{tikzcd}
  \]
  is hence the following commutative diagram:
  \[
    \begin{tikzcd}
        \dotsb
        \arrow{r}
      & P_2
        \arrow{r}[above]{d_2}
      & P_1
        \arrow{r}[above]{d_1}
      & P_0
        \arrow{r}[above]{d_0}
      & P_{-1}
        \arrow{r}[above]{d_{-1}}
        \arrow{d}[right]{\hat{f}_{-1}}
      & P_{-2}
        \arrow{d}[right]{\hat{f}_{-2}}
      \\
        \dotsb
        \arrow{r}
      & C_2
        \arrow{r}[below]{d_2}
      & C_1
        \arrow{r}[below]{d_1}
      & C_0
        \arrow{r}[below]{d_0}
      & C_{-1}
        \arrow{r}[below]{d_{-1}}
      & C_{-2}
    \end{tikzcd}
  \]
  This commutative diagram has exact rows and the object~$P_n$ is projective for every~$n \geq 0$.
  
  We construct the required morphisms~$\hat{f}_n \colon P_n \to C_n$ by induction over~$n \geq -2$, with~$\hat{f}_{-1}$ and~$\hat{f}_{-2}$ being already given.
  So suppose more generally that we have already constructed morphisms~$\hat{f}_n, \dotsc, \hat{f}_0, \hat{f}_{-1}, \hat{f}_{-2}$ for some~$n \geq -1$ that make the diagram
  \[
    \begin{tikzcd}
        P_n
        \arrow{r}[above]{d_n}
        \arrow{d}[right]{\hat{f}_n}
      & P_{n-1}
        \arrow{r}[above]{d_{n-1}}
        \arrow{d}[right]{\hat{f}_{n-1}}
      & \dotsb
        \arrow{r}[above]{d_1}
      & P_0
        \arrow{r}[above]{d_0}
        \arrow{d}[right]{\hat{f}_0}
      & P_{-1}
        \arrow{r}
        \arrow{d}[right]{\hat{f}_{-1}}
      & P_{-2}
        \arrow{d}[right]{\hat{f}_{-2}}
      \\
        C_n
        \arrow{r}[below]{d_n}
      & C_{n-1}
        \arrow{r}[below]{d_{n-1}}
      & \dotsb
        \arrow{r}[below]{d_1}
      & C_0
        \arrow{r}[below]{d_0}
      & C_{-1}
        \arrow{r}
      & C_{-2}
    \end{tikzcd}
  \]
  commute.
  It follows from~$d^P_n d^P_{n+1} = 0$ and~$d^C_n d^C_{n+1} = 0$ that the differentials~$d^P$ and~$d^C$ induce morphisms~$p \colon P_{n+1} \to \ker(d^P_n)$ and~$q \colon C_{n+1} \to \ker(d^C_n)$ that make the triangles
  \[
    \begin{tikzcd}[column sep = tiny]
        P_{n+1}
        \arrow{rr}[above]{d_{n+1}}
        \arrow[dashed]{dr}[below left]{p}
      & {}
      & P_n
      \\
        {}
      & \ker(d_n)
        \arrow{ur}[below right]{i}
      & {}
    \end{tikzcd}
    \qquad\text{and}\qquad
    \begin{tikzcd}[column sep = tiny]
        C_{n+1}
        \arrow{rr}[above]{d_{n+1}}
        \arrow[dashed]{dr}[below left]{q}
      & {}
      & C_n
      \\
        {}
      & \ker(d_n)
        \arrow{ur}[below right]{j}
      & {}
    \end{tikzcd}
  \]
  commute.
  It follows from the commutativity of the square
  \[
    \begin{tikzcd}
        P_n
        \arrow{r}[above]{d_n}
        \arrow{d}[left]{\hat{f}_n}
      & P_{n-1}
        \arrow{d}[right]{\hat{f}_{n-1}}
      \\
        C_n
        \arrow{r}[above]{d_n}
      & C_{n-1}
    \end{tikzcd}
  \]
  that the morphism~$\hat{f}_n \colon P_n \to C_n$ induces a morphism~$\tilde{f}_n \colon \ker(d^P_n) \to \ker(d^C_n)$ that makes the following diagram commute:
  \[
    \begin{tikzcd}[column sep = small]
        P_{n+1}
        \arrow{rrr}[above, near start]{d_{n+1}}
        \arrow{drr}[below left]{p}
      & {}
      & {}
      & P_n
        \arrow{dd}[right]{\hat{f}_n}
      \\
        {}
      & {}
      & \ker(d_n)
        \arrow{ur}[below right]{i}
      & {}
      \\
        C_{n+1}
        \arrow{rrr}[above, near start]{d_{n+1}}
        \arrow{drr}[below left]{q}
      & {}
      & {}
      & C_n
      \\
        {}
      & {}
      & \ker(d_n)
        \arrow[dashed, from=uu, crossing over, "\tilde{f}_n", right, near start]
        \arrow{ur}[below right]{j}
      & {}
    \end{tikzcd}
  \]
  The morphism~$q \colon C_{n+1} \to \ker(d^C_n)$ is an epimorphism because it is the composition
  \[
    q
    \colon
    C_{n+1}
    \to
    \im(d^C_{n+1})
    \to
    \ker(d^C_n)
  \]
  with~$C_{n+1} \to \im(d_{n+1})$ being an epimorphism, and~$\im(d^C_{n+1}) \to \ker(d^C_n)$ being an isomorphism by the exactness of the sequence~$C_{n+1} \xto{d_{n+1}} C_n \xto{d_n} C_{n-1}$.
  It follows from~$P_{n+1}$ being projective that there exist a morphism~$\hat{f}_{n+1} \colon P_{n+1} \to C_{n+1}$ that makes the square
  \[
    \begin{tikzcd}
        P_{n+1}
        \arrow{r}[above]{p}
        \arrow[dashed]{d}[left]{\hat{f}_{n+1}}
      & \ker(d_n)
        \arrow{d}[right]{\tilde{f}_n}
      \\
        C_{n+1}
        \arrow{r}[below]{q}
      & \ker(d_n)
    \end{tikzcd}
  \]
  commute.
  Then also the square
  \[
    \begin{tikzcd}[column sep = large]
        P_{n+1}
        \arrow{r}[above]{d_{n+1}}
        \arrow[dashed]{d}[left]{\hat{f}_{n+1}}
      & P_n
        \arrow{d}[right]{\hat{f}_n}
      \\
        C_{n+1}
        \arrow{r}[above]{d_n}
      & C_n
    \end{tikzcd}
  \]
  commute, i.e.\ the complete diagram
  \[
    \begin{tikzcd}[column sep = small]
        P_{n+1}
        \arrow{rrr}[above, near start]{d_{n+1}}
        \arrow{drr}[below left]{p}
        \arrow[dashed]{dd}[left]{\hat{f}_{n+1}}
      & {}
      & {}
      & P_n
        \arrow{dd}[right]{\hat{f}_n}
      \\
        {}
      & {}
      & \ker(d_n)
        \arrow{ur}[below right]{i}
      & {}
      \\
        C_{n+1}
        \arrow{rrr}[above, near start]{d_{n+1}}
        \arrow{drr}[below left]{q}
      & {}
      & {}
      & C_n
      \\
        {}
      & {}
      & \ker(d_n)
        \arrow[from=uu, crossing over, "\tilde{f}_n", right, near start]
        \arrow{ur}[below right]{j}
      & {}
    \end{tikzcd}
  \]
  commutes.
  Indeed, we have that
  \[
      d_{n+1} \hat{f}_{n+1}
    = j q \hat{f}_{n+1}
    = j \tilde{f}_n p
    = \hat{f}_n i p
    = \hat{f}_n d_{n+1} \,.
  \]
  
  To show that the morphism~$\hat{f}$ is unique up to homotopy we may assume that~$f = 0$:
  Indeed, if~$\hat{f}' \colon \Pcc \to \Ccc$ is another lift of~$f \colon X \to Y$ then the difference~$\hat{f}' - \hat{f}$ is a lift of~$f - f = 0$.
  And the morphisms~$\hat{f}'$ and~$\hat{f}$ are homotopic if and only if this difference~$\hat{f}' - \hat{f}$ is null homotopic, i.e.\ is homotopic to the zero morphism~$\Pcc \to \Ccc$, which is another lift of the zero morphism~$X \to Y$.
  
  So let~$f = 0$, and let us shows that any lift~$\hat{f} \colon \Pcc \to \Ccc$ is null homotopic.
  For this we need to construct morphisms~$s_n \colon P_n \to C_{n+1}$ for~$n \geq 0$ such that
  \[
    d_1 s_0 = \hat{f}_0
    \qquad\text{and}\qquad
    \text{$d_{n+1} s_n + s_{n-1} d_n = \hat{f}_n$ for every~$n \geq 1$} \,.
  \]
  We may visualize the role of the morphisms~$s_n$ as follows:
  \[
    \begin{tikzcd}[row sep = large]
        \dotsb
        \arrow{r}
      & P_2
        \arrow{r}[above]{d_2}
        \arrow{d}[right, near start]{\hat{f}_2}
        \arrow[dashed]{dl}
      & P_1
        \arrow{r}[above]{d_1}
        \arrow{d}[right, near start]{\hat{f}_1}
        \arrow[dashed]{dl}[below right]{s_1}
      & P_0
        \arrow{r}[above]{d_0}
        \arrow{d}[right]{\hat{f}_0}
        \arrow[dashed]{dl}[below right]{s_0}
      & P_{-1}
        \arrow{r}{d_{-1}}
        \arrow{d}[right]{0}
      & P_{-2}
        \arrow{d}[right]{0}
      \\
        \dotsb
        \arrow{r}
      & C_2
        \arrow{r}[below]{d_2}
      & C_1
        \arrow{r}[below]{d_1}
      & C_0
        \arrow{r}[below]{d_0}
      & C_{-1}
        \arrow{r}[below]{d_{-1}}
      & C_{-2}
    \end{tikzcd}
  \]
  We define~$s_{-1} \colon P_{-1} \to C_0$ and~$s_{-2} \colon P_{-2} \to C_{-1}$ as the respective zero morphisms, and hence need to show that
  \[
      \hat{f}_n
    = d_{n+1} s_n + s_{n-1} d_n
  \]
  for every~$n \geq -1$, with this relation already holding for~$n = -1$.
  
  Suppose that~$s_{n-1}, \dotsc, s_{-1}, s_{-2}$ are already constructed for some~$n \geq 0$.
  Then the morphism
  \[
    \hat{f}_n - s_{n-1} d_n
    \colon
    P_n
    \to
    C_n
  \]
  satisfies
  \begin{align*}
        d_n ( \hat{f}_n - s_{n-1} d_n)
    &=  d_n \hat{f}_n - d_n s_{n-1} d_n \\
    &= d_n \hat{f}_n - ( \hat{f}_{n-1} - s_{n-2} d_{n-1} ) d_n  \\
    &= d_n \hat{f}_n - \hat{f}_{n-1} d_n  \\
    &= 0
  \end{align*}
  because~$\hat{f}$ is a morphism of chain complexes.
  It follows from the universal property of the kernel~$\ker(d_n) \to C_n$ that there exists a (unique) morphism~$\tilde{s}_n \colon P_n \to \ker(d_n)$ that makes the diagram
  \[
    \begin{tikzcd}
        {}
      & P_n
        \arrow[bend left]{drr}[above right]{0}
        \arrow{d}[right]{\hat{f}_n - s_{n-1} d_n}
        \arrow[dashed]{dl}[above left]{\tilde{s}_n}
      & {}
      \\
        \ker(d_n)
        \arrow{r}
      & C_n
        \arrow{rr}[below]{d_n}
      & {}
      & C_{n+1}
    \end{tikzcd}
  \]
  commute.
  The canonical morphism~$C_{n+1} \to \ker(d_n)$ that makes the diagram
  \[
    \begin{tikzcd}
        C_{n+1}
        \arrow[dashed]{r}
        \arrow[bend right]{rr}[below]{d_{n+1}}
      & \ker(d_n)
        \arrow{r}
      & C_n
    \end{tikzcd}
  \]
  commute is an epimorphism by the exactness of the sequence~$C_{n+1} \to C_n \to C_{n-1}$, as seen in the previous part of the proof.
  It follows from~$P_n$ being projective that there exists a morphism~$s_n \colon P_n \to C_{n+1}$ that makes the diagram
  \[
    \begin{tikzcd}
        {}
      & {}
      & P_n
        \arrow{d}[right]{\hat{f}_n - s_{n-1} d_n}
        \arrow{dl}[above left]{\tilde{s}_n}
        \arrow[dashed, bend right]{dll}[above left]{s_n}
      \\
        C_{n+1}
        \arrow{r}
        \arrow[bend right]{rr}[below]{d_{n+1}}
      & \ker(d_n)
        \arrow{r}
      & C_n
    \end{tikzcd}
  \]
  commute.
  Then
  \[
      d_{n+1} s_n + s_{n-1} d_n
    = \hat{f}_n - s_{n-1} d_n + s_{n-1} d_n
    = \hat{f}_n \,,
  \]
  as desired.
\end{proof}


\begin{remark*}
  One may rewrite the commutative square~\eqref{comparison square for projective} as
  \[
    \begin{tikzcd}
        \Pcc
        \arrow{r}[above]{p_0}
        \arrow[dashed]{d}[left]{\hat{f}}
      & X
        \arrow{d}[right]{f}
      \\
        \Ccc
        \arrow{r}[above]{c_0}
      & Y
    \end{tikzcd}
  \]
  and the commutative square~\eqref{comparison square for injective} as follows:
  \[
    \begin{tikzcd}
        X
        \arrow{r}[above]{c^0}
        \arrow{d}[left]{f}
      & \Cccc
        \arrow[dashed]{d}[right]{\hat{f}}
      \\
        Y
        \arrow{r}[above]{i^0}
      & \Iccc
    \end{tikzcd}
  \]
\end{remark*}


\begin{lemma}[Horseshoe lemma]\index{horseshoe lemma}\index{lemma!horseshoe}
  \leavevmode
  \begin{enumerate}
    \item
      \label{projective horseshoe}
      Let~$0 \to X' \to X \to X'' \to 0$ be a short exact sequence in~$\Acat$.
      If~$\Pcc' \xto{p'_0} X'$ and~$\Pcc'' \xto{p''_0} X''$ are projective resolutions then there exists a projective resolution~$\Pcc \xto{p_0} X$ of~$X$ such that the short exact sequence~$0 \to X' \to X \to X'' \to 0$ lifts to a short exact sequence of chain complexes
      \[
        0
        \to
        \Pcc'
        \to
        \Pcc
        \to
        \Pcc''
        \to
        0 \,,
      \]
      in the sense that the following diagram commutes: 
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & \Pcc'
            \arrow{r}
            \arrow{d}[right]{p'_0}
          & \Pcc
            \arrow{r}
            \arrow{d}[right]{p_0}
          & \Pcc''
            \arrow{r}
            \arrow{d}[right]{p''_0}
          & 0
          \\
            0
            \arrow{r}
          & X'
            \arrow{r}
          & X
            \arrow{r}
          & X''
            \arrow{r}
          & 0
        \end{tikzcd}
      \]
      In other words, the following diagram commutes:
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & P'_0
            \arrow{r}
            \arrow{d}[right]{p'_0}
          & P_0
            \arrow{r}
            \arrow{d}[right]{p_0}
          & P''_0
            \arrow{r}
            \arrow{d}[right]{p''_0}
          & 0
          \\
            0
            \arrow{r}
          & X'
            \arrow{r}
          & X
            \arrow{r}
          & X''
            \arrow{r}
          & 0
        \end{tikzcd}
      \]
    \item
      \label{injective horseshoe}
      Let~$0 \to \bp X \to X \to \bpp X \to 0$ be a short exact sequence in~$\Acat$.
      If~$\bp X \xto{\bp i^0} \bp\Iccc$ and~$\bpp X \xto{\bpp i^0} \bpp\Iccc$ are injective resolutions then there exists an injective resolution~$X \xto{i^0} \Iccc$ of~$X$ such that the short exact sequence~$0 \to \bp X \to X \to \bpp X \to 0$ lifts to a short exact sequence of cochain complexes
      \[
        0
        \to
        \bp\Iccc
        \to
        \Iccc
        \to
        \bpp\Iccc
        \to
        0 \,,
      \]
      in the sense that the following diagram commutes:
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & \bp X
            \arrow{r}
            \arrow{d}[right]{\bp i^0}
          & X
            \arrow{r}
            \arrow{d}[right]{i^0}
          & \bpp X
            \arrow{r}
            \arrow{d}[right]{\bpp i^0}
          & 0
          \\
            0
            \arrow{r}
          & \bp\Iccc
            \arrow{r}
          & \Iccc
            \arrow{r}
          & \bpp\Iccc
            \arrow{r}
          & 0
        \end{tikzcd}
      \]
      In other words, the following diagram commute:
      \[
        \begin{tikzcd}
            0
            \arrow{r}
          & \bp X
            \arrow{r}
            \arrow{d}[right]{\bp i^0}
          & X
            \arrow{r}
            \arrow{d}[right]{i^0}
          & \bpp X
            \arrow{r}
            \arrow{d}[right]{\bpp i^0}
          & 0
          \\
            0
            \arrow{r}
          & \bp I^0
            \arrow{r}
          & I^0
            \arrow{r}
          & \bpp I^0
            \arrow{r}
          & 0
        \end{tikzcd}
      \]
  \end{enumerate}
\end{lemma}


\begin{proof}
  It sufficies to prove part~\ref*{projective horseshoe}, as part~\ref*{injective horseshoe} follows from part~\ref*{projective horseshoe} by duality.
  We denote the given short exact sequence by
  \[
    0
    \to
    X'
    \xlongto{f}
    X
    \xlongto{g}
    X''
    \to
    0 \,.
  \]
  
  We set~$P_n \defined P'_n \oplus P''_n$ for every~$n \geq 0$
  and consider for every~$n \geq 0$ the short exact sequence
  \[
    0
    \to
    P'_n
    \xlongto{ \begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix} }
    P_n
    \xlongto{ \begin{bsmallmatrix} 0 & \id \end{bsmallmatrix} }
    P''_n
    \to
    0
  \]
  We need to construct morphisms~$d_n \colon P_n \to P_{n-1}$ and~$p_0 \colon P_0 \to X$ that make the diagram
  \[
    \begin{tikzcd}[ampersand replacement = \&, column sep = large]
          {}
      \&  \vdots
          \arrow{d}
      \&  \vdots
          \arrow[dashed]{d}
      \&  \vdots
          \arrow{d}
      \&  {}
      \\
          0
          \arrow{r}
      \&  P'_2
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[left]{d'_2}
      \&  P_2
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow[dashed]{d}[right]{d_2}
      \&  P''_2
          \arrow{r}
          \arrow{d}[right]{d''_2}
      \&  0
      \\
          0
          \arrow{r}
      \&  P'_1
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[left]{d'_1}
      \&  P_1
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow[dashed]{d}[right]{d_1}
      \&  P''_1
          \arrow{r}
          \arrow{d}[right]{d''_1}
      \&  0
      \\
          0
          \arrow{r}
      \&  P'_0
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[left]{p'_0}
      \&  P_0
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow[dashed]{d}{p_0}
      \&  P''_0
          \arrow{r}
          \arrow{d}[right]{p''_0}
      \&  0
      \\
          0
          \arrow{r}
      \&  X'
          \arrow{r}[above]{f}
          \arrow{d}
      \&  X
          \arrow{r}[above]{g}
          \arrow{d}
      \&  X''
          \arrow{r}
          \arrow{d}
      \&  0
      \\
          {}
      \&  0
      \&  0
      \&  0
      \&  {}
    \end{tikzcd}
  \]
  commute and that make the middle column into a projective resolution of~$X$.
  That the sequence~$0 \to \Pcc' \to \Pcc \to \Pcc'' \to 0$ of chain complexes is then already (short) exact follows because exactness in~$\Ch(\Acat)$ is component componentwise (as explained in \cref{exactness for chain complexes is computed degreewise}).
  
  We start by contructing the morphism~$p_0$:
  We have the following diagram in which the morphism~$g$ is an epimorphism and the object~$P''_0$ is projective:
  \[
    \begin{tikzcd}
        {}
      & P'_0
        \arrow{d}[right]{p'_0}
      & {}
      & P''_0
        \arrow{d}[right]{p''_0}
      & {}
      \\
        0
        \arrow{r}
      & X
        \arrow{r}[above]{f}
        \arrow{d}
      & Y
        \arrow{r}[above]{g}
      & Z 
        \arrow{r}
        \arrow{d}
      & 0
      \\
        {}
      & 0
      & {}
      & 0
      & {}
    \end{tikzcd}
  \]
  It follows that there exist a morphism~$g' \colon P''_0 \to X$ that makes the triangle
  \[
    \begin{tikzcd}
        {}
      & P''_0
        \arrow{d}[right]{p''_0}
        \arrow[dashed]{dl}[above left]{g'}
      \\
        X
        \arrow{r}[below]{g}
      & X''
    \end{tikzcd}
  \]
  commute.
  We define
  \[
    p_0
    \defined
    \begin{bmatrix}
      f p'_0  & g'
    \end{bmatrix}
    \colon
    P_0
    =
    P'_0 \oplus P''_0
    \to
    X \,.
  \]
  Then the diagram
  \[
    \begin{tikzcd}[ampersand replacement = \&]
          0
          \arrow{r}
      \&  P'_0
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[left]{p'_0}
      \&  P_0
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow[dashed]{d}[right]{p_0}
      \&  P''_0
          \arrow{r}
          \arrow{d}[right]{p''_0}
      \&  0
      \\
          0
          \arrow{r}
      \&  X'
          \arrow{r}[above]{f}
      \&  X
          \arrow{r}[above]{g}
      \&  X''
          \arrow{r}
      \&  0
    \end{tikzcd}
  \]
  commutes because
  \begin{gather*}
      p_0
      \begin{bmatrix}
        \id \\
        0
      \end{bmatrix}
    = \begin{bmatrix}
        f p'_0  & g'
      \end{bmatrix}
      \begin{bmatrix}
        \id \\
        0
      \end{bmatrix}
    = f p'_0
  \shortintertext{and}
      g p_0
    = g
      \begin{bmatrix}
        f p'_0  & g'
      \end{bmatrix}
    = \begin{bmatrix}
        g f p'_0  & g g'
      \end{bmatrix}
    = \begin{bmatrix}
        0 & p''_0
      \end{bmatrix}
    = p''_0
      \begin{bmatrix}
        0 & \id
      \end{bmatrix} \,.
  \end{gather*}
  It follows from the~\hyperref[5 lemma]{\dash{5}{lemma}} that the morphisms~$p_0$ is an epimorphism, because both~$p'_0$ and~$p''_0$ are epimorphism.
  We have hence the following commutative diagram with exact column:
  \[
    \begin{tikzcd}[ampersand replacement = \&, column sep = large]
          {}
      \&  \vdots
          \arrow{d}
      \&  {}
      \&  \vdots
          \arrow{d}
      \&  {}
      \\
          {}
      \&  P'_2
          \arrow{d}[right]{d'_2}
      \&  {}
      \&  P''_2
          \arrow{d}[right]{d''_2}
      \&  {}
      \\
          {}
      \&  P'_1
          \arrow{d}[right]{d'_1}
      \&  {}
      \&  P''_1
          \arrow{d}[right]{d''_1}
      \&  {}
      \\
          0
          \arrow{r}
      \&  P'_0
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[right]{p'_0}
      \&  P_0
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow{d}[right]{p_0}
      \&  P''_0
          \arrow{r}
          \arrow{d}[right]{p''_0}
      \&  0
      \\
          0
          \arrow{r}
      \&  X'
          \arrow{r}[below]{f}
          \arrow{d}
      \&  X
          \arrow{r}[below]{g}
          \arrow{d}
      \&  X''
          \arrow{r}
          \arrow{d}
      \&  0
      \\
          {}
      \&  0
      \&  0
      \&  0
      \&  {}
    \end{tikzcd}
  \]
  
  To construct the morphism~$d_1 \colon P_1 \to P_0$ we proceed similar as before:
  We get the following commutative diagram with exact columns:
  \[
    \begin{tikzcd}[ampersand replacement = \&]
          {}
      \&  0
          \arrow{d}
      \&  0
          \arrow{d}
      \&  0
          \arrow{d}
      \&  {}
      \\
          0
          \arrow{r}
      \&  \ker(p'_0)
          \arrow{r}
          \arrow{d}[right]{\tilde{d}'_0}
      \&  \ker(p_0)
          \arrow{r}
          \arrow{d}[right]{\tilde{d}_0}
      \&  \ker(p''_0)
          \arrow{r}
          \arrow{d}[right]{\tilde{d}''_0}
      \& 0
      \\
          0
          \arrow{r}
      \&  P'_0
          \arrow{r}
          \arrow{d}[right]{p'_0}
      \&  P_0
          \arrow{r}
          \arrow{d}[right]{p_0}
      \&  P''_0
          \arrow{r}
          \arrow{d}[right]{p''_0}
      \&  0
      \\
          0
          \arrow{r}
      \&  X'
          \arrow{r}[below]{f}
          \arrow{d}
      \&  X
          \arrow{r}[below]{g}
          \arrow{d}
      \&  X''
          \arrow{r}
          \arrow{d}
      \&  0
      \\
          {}
      \&  0
      \&  0
      \&  0
      \&  {}
    \end{tikzcd}
  \]
  The lower two rows are exact, hence the upper row is also exact by the~\hyperref[9 lemma]{\dash{9}{lemma}}.
  We can therefore consider the following subdiagram, in which the morphism~$\ker(p_0) \to \ker(p''_0)$ is an epimorphism and the object~$P''_1$ is projective.
  \[
    \begin{tikzcd}
        {}
      & P'_1
        \arrow{d}[right]{\tilde{d}'_1}
      & {}
      & P''_1
        \arrow{d}[right]{\tilde{d}''_1}
      & {}
      \\
        0
        \arrow{r}
      & \ker(p'_0)
        \arrow{r}
        \arrow{d}
      & \ker(p_0)
        \arrow{r}
      & \ker(p''_0)
        \arrow{r}
        \arrow{d}
      & 0
      \\
        {}
      & 0
      & {}
      & 0
      & {}
    \end{tikzcd}
  \]
  We find as before that there exists an epimorphism~$d_1 \colon P_1 \to \ker(p_0)$ that makes the diagram
  \[
    \begin{tikzcd}[ampersand replacement = \&]
          0
          \arrow{r}
      \&  P'_1
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[right]{\tilde{d}'_1}
      \&  P_1
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow[dashed]{d}[right]{\tilde{d}_1}
      \&  P''_1
          \arrow{r}
          \arrow{d}[right]{\tilde{d}''_1}
      \&  0
      \\
          0
          \arrow{r}
      \&  \ker(p'_0)
          \arrow{r}
      \&  \ker(p_0)
          \arrow{r}
      \&  \ker(p''_0)
          \arrow{r}
      \&  0
    \end{tikzcd}
  \]
  commute.
  It follows from the exactness of the sequence
  \[
    \ker(p_0)
    \xlongto{\tilde{d}_0}
    P_0
    \xlongto{p_0}
    X
  \]
  that the sequence
  \[
    P_1
    \xlongto{\tilde{d}_0 \tilde{d}_1}
    P_0
    \xlongto{p_0}
    X
  \]
  is again exact because~$\tilde{d}_1$ is an epimorphism.
  We have hence arrived at the following commutatitve diagram with exact columns:
  \[
    \begin{tikzcd}[ampersand replacement = \&, column sep = large]
          {}
      \&  \vdots
          \arrow{d}
      \&  {}
      \&  \vdots
          \arrow{d}
      \&  {}
      \\
          {}
      \&  P'_2
          \arrow{d}[right]{d'_2}
      \&  {}
      \&  P''_2
          \arrow{d}[right]{d''_2}
      \&  {}
      \\
          0
          \arrow{r}
      \&  P'_1
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[right]{d'_1}
      \&  P_1
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow{d}[right]{d_0}
      \&  P''_1
          \arrow{r}
          \arrow{d}[right]{d''_1}
      \& 0
      \\
          0
          \arrow{r}
      \&  P'_0
          \arrow{r}[above]{\begin{bsmallmatrix} \id \\ 0 \end{bsmallmatrix}}
          \arrow{d}[right]{p'_0}
      \&  P_0
          \arrow{r}[above]{\begin{bsmallmatrix} 0 & \id \end{bsmallmatrix}}
          \arrow{d}[right]{p_0}
      \&  P''_0
          \arrow{r}
          \arrow{d}[right]{p''_0}
      \&  0
      \\
          0
          \arrow{r}
      \&  X'
          \arrow{r}[below]{f}
          \arrow{d}
      \&  X
          \arrow{r}[below]{g}
          \arrow{d}
      \&  X''
          \arrow{r}
          \arrow{d}
      \&  0
      \\
          {}
      \&  0
      \&  0
      \&  0
      \&  {}
    \end{tikzcd}
  \]
  We can now continue by induction.
\end{proof}


\begin{remark*}
  The term \enquote{horse shoe lemma} comes from the form of the following diagram, that we are filling in with an exact column in the middle:
  \[
    \begin{tikzcd}[ampersand replacement = \&, column sep = large]
          {}
      \&  \vdots
          \arrow{d}
      \&  {}
      \&  \vdots
          \arrow{d}
      \&  {}
      \\
          {}
      \&  P'_2
          \arrow{d}[left]{d'_2}
      \&  {}
      \&  P''_2
          \arrow{d}[right]{d''_2}
      \&  {}
      \\
          {}
      \&  P'_1
          \arrow{d}[left]{d'_1}
      \&  {}
      \&  P''_1
          \arrow{d}[right]{d''_1}
      \&  {}
      \\
          {}
      \&  P'_0
          \arrow{d}[left]{p'_0}
      \&  {}
      \&  P''_0
          \arrow{d}[right]{p''_0}
      \&  {}
      \\
          0
          \arrow{r}
      \&  X'
          \arrow{r}[above]{f}
          \arrow{d}
      \&  X
          \arrow{r}[above]{g}
          \arrow{d}
      \&  X''
          \arrow{r}
          \arrow{d}
      \&  0
      \\
          {}
      \&  0
      \&  0
      \&  0
      \&  {}
    \end{tikzcd}
  \]
\end{remark*}





