\section{Special Objects}


\begin{definition}
  Let~$X$ be an object in a category~$\Ccat$.
  \begin{enumerate}
    \item
      The object~$X$ is \emph{initial}\index{initial object}\index{object!initial} if there exists  for every object~$Y \in \Ob(\Ccat)$ a unique morphism~$X \to Y$ in~$\Ccat$.
    \item
      The object~$X$ is \emph{terminal}\index{terminal object}\index{object!terminal} or \emph{final}\index{final object}\index{object!final} if there exists for every object~$Y \in \Ob(\Ccat)$ a unique morphism~$Y \to X$ in~$\Ccat$.
    \item
      The object~$X$ is a \emph{zero object}\index{zero!object}\index{object!zero} if it is both initial and terminal.
  \end{enumerate}
\end{definition}


\begin{remark}
  Let~$\Ccat$ be a category.
  \begin{enumerate}
    \item
      An object~$X$ of~$\Ccat$ is initial (in~$\Ccat$) if and only if it is terminal in~$\Ccat^\op$.
    \item
      Initial and terminal objects are unique up to unique isomorphisms (if they exist).
    \item
      If~$\Ccat$ admits a zero object then it is denoted by~$0 = 0_\Ccat$.
  \end{enumerate}
\end{remark}


\begin{example}
  \leavevmode
  \begin{enumerate}
    \item
      In the category~$\Set$, the empty set~$\emptyset$ is the unique initial object, and every \dash{one}{point} set~$\{\ast\}$ is a final object.
      The analogous statements hold for the category~$\Top$.
    \item
      In the category~$\Modl{A}$ the zero module~$0$ is the zero object.
    \item
      In the category~$\Group$ the trivial groups~$1$ is the zero object.
    \item
      In the category~$\kAlg$ the~{\kalg}~$\kf$ is initial, while the zero algebra~$0$ is final.
  \end{enumerate}
\end{example}


\begin{example*}
  The category~$\Field$ of fields has neither an intial nor a final object.
% TODO: Explain this.
\end{example*}


\begin{remarkdefinition}
  Let~$\Ccat$ be a category that admits a zero object~$0$.
  Then there exists for any two objects~$X, Y \in \Ob(\Ccat)$ a unique morphism~$0_{XY} \colon X \to Y$ that factors trough the zero object, i.e.\ that makes the triangle
  \[
    \begin{tikzcd}
        X
        \arrow{dr}
        \arrow{rr}[above]{0_{XY}}
      & {}
      & Y
      \\
        {}
      & 0
        \arrow{ur}
      & {}
    \end{tikzcd}
  \]
  commute.
  The morphism~$0_{XY}$ is the \emph{zero morphism}\index{zero!morphism}\index{morphism!zero} from~$X$ to~$Y$.
\end{remarkdefinition}


\begin{remark*}
  Let~$\Ccat$ be a category that admits a zero object~$0$.
  Then it holds for every morphism~$f \colon X \to Y$ that
  \[
      f \circ 0_{W,X}
    = 0_{W,Y}
    \qquad\text{and}\qquad
      0_{Y,Z} \circ f
    = 0_{X,Z}
  \]
  for all~$W, Z \in \Ob(\Ccat)$.
  Indeed, we have the following commutative diagrams:
  \[
    \begin{tikzcd}
        W
        \arrow[dashed, bend left = 40]{rrr}[above]{f \circ 0_{W,X}}
        \arrow{rr}[above]{0_{W,X}}
        \arrow{dr}
      & {}
      & X
        \arrow{r}[above]{f}
      & Y
      \\
        {}
      & 0
        \arrow{ur}
        \arrow[dashed, bend right]{urr}
      & {}
      & {}
    \end{tikzcd}
    \qquad
    \begin{tikzcd}
        X
        \arrow[dashed, bend left = 40]{rrr}[above]{0_{Y,Z} \circ f}
        \arrow{r}[above]{f}
        \arrow[dashed, bend right]{drr}
      & Y
        \arrow{rr}[above]{0_{Y,Z}}
        \arrow{dr}
      & {}
      & Z
      \\
        {}
      & {}
      & 0
        \arrow{ur}
      & {}
    \end{tikzcd}
  \]
  The commutativity of the first diagram shows that~$f \circ 0_{W,X}$ factors through the zero object~$0$, and is hence the zero morphism~$W \to Y$.
  The commutativity of the second diagram shows similarly that~$0_{Y,Z} \circ f$ factors through the zero objects~$0$ and is therefore the zero morphism~$X \to Z$.
\end{remark*}





\lecturend{8}




